CrystalPlan: an Experiment Planning Tool for Crystallography

### 2010-09-24 docs/CrystalPlan_paper/paper.tex
Refs #182: Xiaoping's revisions and new figures for CrystalPlan paper.

\begin{abstract}
Beam time at large x-ray and neutron scattering facilities is always at a premium.
The CrystalPlan program can calculate the data coverage of a crystal in reciprocal
space in a single-crystal diffraction time-of-flight experiment. CrystalPlan can 
help a user build an experiment plan that will acquire the most data possible, 
with sufficient coverage but limited redundancy, therefore increasing scientific 
productivity. 
An attractive GUI including a 3D viewer and an automated coverage optimizer 
are among its useful features. 
A sample use case of the program with the TOPAZ beamline at SNS will be
presented. 
\end{abstract}

@@

The TOPAZ beamline at the Spallation Neutron Source (SNS) at Oak Ridge National
Lab (ORNL), currently finishing commissioning, is a neutron time-of-flight
instrument designed to acquire single-crystal diffraction data of
small-to-moderate sized unit cells with high throughput. The final design calls
for an array of 48 detectors covering a large fraction of  $2\pi$ steradians;
however as of this writing, only 14 of the 48 detectors are installed. In order
to have sufficient data to perform a fit, an experimenter will  typically want
to measure $ > 85\%$ of the peaks within a certain range of d-spacings. 
However, TOPAZ's complex geometry makes it very difficult to come up with an
effective set of sample orientations that will cover all the required peaks
without unnecessary redundancy.

Existing diffraction experiment planning software programs (for example,
CrysAlis Pro from Oxford Diffraction; DIFFRAC from Bruker AXS) were designed for
X-ray experiments where a single  incident wavelength and a large number of
sample orientations or sweeps are used to collect data. 
These programs cannot be used for TOPAZ, since as a time-of-flight instrument, 
it is able to collect data from a wide bandwidth of incident neutrons. 
Additionally, this means that relatively few sample orientations are needed in
order to complete a measurement: whereas an X-ray measurement might have
hundreds or thousands of individual frames, a TOPAZ run may have as few as 10.

@@

The resulting UB matrix was loaded into CrystalPlan, where the Automatic
Coverage Optimizer was used to generate an experiment plan with 8 sample
orientations that covered 95\% of the peaks, taking into account the crystal
symmetry (orthorombic or mmm). The plan was left to run unattended overnight.

Peaks were integrated using ISAW, and the resulting .integrate file was loaded
into CrystalPlan for comparison. The predicted and actual measurement
positions on each detector face were compared for the 554 integrated peaks
found. Very good agreement was found; the
RMS difference between the predicted and actual measurement wavelength was only
0.016 \ang, whereas the RMS of the detector position error was
7.6 mm. Inspecting the offsets showed that the main contributor to this error was a
slight misalignment of the goniometer, which CrystalPlan can help diagnose.

@@

>>>>

\begin{abstract}
Beam time at large x-ray and neutron scattering facilities is always at a premium.
The CrystalPlan program can calculate the data coverage of a crystal in reciprocal
space in a single-crystal diffraction time-of-flight experiment. CrystalPlan can 
help a user build an experiment plan that will acquire the most data possible, 
with sufficient coverage but limited redundancy, therefore increasing scientific 
productivity. 
A user friendly GUI including a 3D viewer, an automated coverage optimizer,
and an option to reorient the crystal for the measurement 
of selected hkls on specific detector positions 
are among its useful features. 
A sample use case of the program with the TOPAZ beamline at SNS will be
presented. 
\end{abstract}

@@

The TOPAZ beamline at the Spallation Neutron Source (SNS) at Oak Ridge National
Lab (ORNL), currently finishing commissioning, is a neutron time-of-flight
Laue instrument with high wavelength resolution. It is designed to acquire
single-crystal diffraction data of small-to-moderate sized unit cells with  high
throughput. The final design calls for an array of 48 detectors covering a large
fraction of $2\pi$ steradians. However as of this writing, only 14 of the 48
detectors are installed, giving only a fraction of $2\pi$ steradian coverage. 
In order to have sufficient data to perform a fit, an experimenter will typically want
to measure $ > 85\%$ of the peaks within a certain range of d-spacings. 
However, wide wavelength band combined with TOPAZ's complex
geometry makes it very difficult to come up with an effective set of sample
orientations that will cover all the required peaks without unnecessary redundancy.

Existing diffraction experiment planning software programs (for example,
CrysAlis Pro from Oxford Diffraction; COSMO from Bruker AXS) were designed for
X-ray experiments where a single  incident wavelength and a large number of
sample orientations or sweeps are used to collect data. 
These programs are not suitable for TOPAZ, since as a time-of-flight instrument, 
it is able to collect data from a wide bandwidth of
incident neutrons, using a stationary crystal. 
Additionally, this means that relatively few sample orientations are needed in
order to complete a measurement: whereas an X-ray measurement might have
hundreds or thousands of individual frames, a TOPAZ run may have as few as 10.

@@

The resulting UB matrix was loaded into CrystalPlan, where the Automatic
Coverage Optimizer was used to generate an experiment plan with 8 sample
orientations (Figure \ref{fig:natrolite_gui}a) that covered 97\% of the peaks,
taking into account the crystal symmetry (orthorombic or mmm) (Figure
\ref{fig:natrolite_gui}b). The plan was left to run unattended overnight.

Peaks were integrated using ISAW, and the resulting .integrate file was loaded
into CrystalPlan for comparison. The predicted and actual measurement
positions on each detector face were compared for the 554 integrated peaks
found. Very good agreement was found; the
RMS difference between the predicted and actual measurement wavelength was only
0.016 \ang. Inspecting the detector position offsets showed that the main
contributor to this error was a slight misalignment of the goniometer, which
CrystalPlan can help diagnose.

%, whereas the RMS of the detector position error was 7.6 mm.

@@

\begin{figure}
\caption{a) Screenshot of the Natrolite experiment plan generated using the
Automatic coverage Optimizer. b) Screenshot of the resulting coverage
statistics.}
\includegraphics[width=3.5in]{Plan8List.eps}
a)
\includegraphics[width=3.5in]{NatCoverage.eps}
b)
\label{fig:natrolite_gui}
\end{figure}
